---
title: "Exploratory Analysis"
author: "Amy Watt"
date: "2019-03-28"
output: 
  html_document: 
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
aids <- read.csv("AIDSdata.csv")
library(tidyverse)
library(ggplot2)
library(plyr)
```

```{r, echo=FALSE}
#Changing data from numerical to text
aids$censor_text <- NA
aids$censor_d_text <- NA
aids$txgrp_text <- NA
aids$strat2_text <- NA
aids$sex_text <- NA
aids$raceth_text <- NA
aids$ivdrug_text <- NA
aids$hemophil_text <- NA

#Adding a new column in dataframe for combined sex/raceth demographics.

aids$demo <- NA
aids$demo_text <- NA

#Adding a new column to split cd4 into factors
aids <- aids %>% 
  mutate(cd4f = ifelse(cd4 <= 50, "under50", 
                             ifelse(cd4 <=100, "50-100",
                                    ifelse(cd4 <= 150, "100-150",
                                           ifelse(cd4 <=200, "150-200", "over200"))))) %>%
  mutate(cd4f = factor(cd4f,
                       levels = c("under50", "50-100", "100-150","150-200", "over200")))
```

```{r, echo=FALSE}
for (elt in 1:length(aids$id)) {
  s = toString(aids$sex[elt])
  r = toString(aids$raceth[elt])
  aids$demo[elt] <- as.numeric(paste (s, r, sep=''))
  
  if (aids$censor[elt] == 1) {
    aids$censor_text[elt] <- 'AIDS/death'
  } else {
    aids$censor_text[elt] <- 'Otherwise'
  }
  
  if (aids$censor_d[elt] == 1) {
    aids$censor_d_text[elt] <- 'Death'
  } else {
    aids$censor_d_text[elt] <- 'Otherwise'
  }
  
  if (aids$txgrp[elt] == 1) {
    aids$txgrp_text[elt] <- 'Placebo'
  } else {
    aids$txgrp_text[elt] <- 'IDV'
  }
  
  if (aids$strat2[elt] == 1) {
    aids$strat2_text[elt] <- 'CD4 > 50'
  } else {
    aids$strat2_text[elt] <- 'CD4 <= 50'
  }
  
  if (aids$sex[elt] == 1) {
    aids$sex_text[elt] <- 'Male'
  } else {
    aids$sex_text[elt] <- 'Female'
  }
  
  if (aids$raceth[elt] == 1) {
    aids$raceth_text[elt] <- 'White Non-Hispanic'
  } else if (aids$raceth[elt] == 2) {
    aids$raceth_text[elt] <- 'Black Non-Hispanic'
  } else if (aids$raceth[elt] == 3) {
    aids$raceth_text[elt] <- 'Hispanic'
  } else if (aids$raceth[elt] == 4) {
    aids$raceth_text[elt] <- 'Asian/Pacific Islander'
  } else if (aids$raceth[elt] == 5) {
    aids$raceth_text[elt] <- 'American Indian/Alaskan Native'
  } else if (aids$raceth[elt] == 6) {
    aids$raceth_text[elt] <- 'Other/Unknown'
  }
  
  if (aids$ivdrug[elt] == 1) {
    aids$ivdrug_text[elt] <- 'Never'
  } else if (aids$ivdrug[elt] == 2) {
    aids$ivdrug_text[elt] <- 'Currently'
  } else if (aids$ivdrug[elt] == 3) {
    aids$ivdrug_text[elt] <- 'Previously'
  }
  
  if (aids$hemophil[elt] == 1) {
    aids$hemophil_text[elt] <- 'Yes'
  } else if (aids$hemophil[elt] == 0) {
    aids$hemophil_text[elt] <- 'No'
  }
  
  aids$demo_text[elt] <- (paste (aids$sex_text[elt], aids$raceth_text[elt], sep=', '))
}
```  

```{r, echo=FALSE}
dfTab <- as.data.frame(table(aids$sex_text))
colnames(dfTab)[1] <- "n"
dfTab$lab <- as.character(100 * dfTab$Freq / sum(dfTab$Freq))

ggplot(aids, aes(x=sex_text, fill=sex_text)) + geom_bar( ) +
  scale_fill_brewer(palette = "Set2") + labs(title = 'Sex Frequency', x = "Sex", y = "Count") + theme(legend.position = "none") + geom_text(data=dfTab, aes(n, Freq, label = Freq), vjust=-0.3, inherit.aes = FALSE)

chisq.test(table(aids$sex_text))
```
From the p-value of 2.2e-16 in a chi-squared test of proportions, there is a statistically significant difference in the proportion of men enrolled in the study to women.


```{r, echo=FALSE}
dfTab <- as.data.frame(table(aids$raceth_text))
colnames(dfTab)[1] <- "n"
dfTab$lab <- as.character(100 * dfTab$Freq / sum(dfTab$Freq))

ggplot(aids, aes(x=raceth_text, fill=raceth_text)) + geom_bar( ) +
  scale_fill_brewer(palette = "Set2") + labs(title = 'Race/Ethnicity Frequency', x = "Race/Ethnicity", y = "Count") +  theme(legend.position = "none") + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + geom_text(data=dfTab, aes(n, Freq, label = Freq), vjust=-0.3, inherit.aes = FALSE)

chisq.test(table(aids$raceth_text))
```
From the p-value of 2.2e-16 in a chi-squared test of proportions, there is a statistically significant difference in the proportion of race/ethnicity groups enrolled in the study.

```{r, echo=FALSE}
dfTab <- as.data.frame(table(aids$demo_text))
colnames(dfTab)[1] <- "n"
dfTab$lab <- as.character(100 * dfTab$Freq / sum(dfTab$Freq))

ggplot(aids, aes(x=demo_text, fill=demo_text)) + geom_bar( ) +
  scale_fill_brewer(palette = "Set3") + labs(title = 'Sex and Race/Ethnicity Demographics', x = "Demographics", y = "Count") + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + theme(legend.position = "none") + geom_text(data=dfTab, aes(n, Freq, label = Freq), vjust=-0.3, inherit.aes = FALSE)

chisq.test(table(aids$demo_text))
```
From the p-value of 2.2e-16 in a chi-squared test of proportions, there is a statistically significant difference in the proportion of gender+race/ethnicity groups enrolled in the study.


```{r, echo=FALSE}
dfTab <- as.data.frame(table(aids$karnof, aids$censor_text))
colnames(dfTab)[1] <- "n"
colnames(dfTab)[2] <- "censor_text"
dfTab$lab <- as.character(100 * dfTab$Freq / sum(dfTab$Freq))

ggplot(dfTab, aes(x=n, y=Freq, fill=censor_text)) + geom_bar(stat='identity') +
  scale_fill_brewer(palette = "Set3") + labs(title = 'Karnofsky Frequency', x = "Karnofsky Scale", y = "Count") +  theme(legend.position = "right") +  theme(legend.title=element_blank()) + geom_text(aes(x=dfTab$n, y=dfTab$Freq, label=dfTab$Freq), position = position_stack(vjust = 0.5))

chisq.test(aids$censor_text, aids$karnof)
```

The p-value of 2.047e-09 indicates that there is a significant difference in the proportion of individuals who develop AIDS or die across the different Karnofsky scale groups.

```{r, echo=FALSE}
ggplot(aids, aes(x=age)) + geom_histogram(breaks=seq(10, 80, by=5), col='black', fill='turquoise') + 
  labs(title="Histogram for Age", x="Age", y="Count") + 
  xlim(c(10,80)) + ylim(c(0,450))
```

```{r, echo=FALSE}
ggplot(aids, aes(x=cd4)) + geom_histogram(breaks=seq(0, 350, by=5), col='black', fill='turquoise') + 
  labs(title="Histogram for cd4", x="cd4", y="Count") + 
  xlim(c(0,350)) + ylim(c(0,60))
```

The largest cd4 in the study is 348. The CDC states that any cd4 under 200 clinically defines HIV, and cd4 over 500 is considered healthy. Initially, when infected with HIV, cd4 count typically drops to around 500 and tends to decline as the disease progresses. The risk of aids double for every 100 cells/mm^3 drop in cd4, but since all patients are well below the 500 cd4 marker, cd4 will be split into 50 cells/mm^3 categories.

```{r, echo=FALSE}
dfTab <- as.data.frame(table(aids$cd4f, aids$censor_text))
colnames(dfTab)[1] <- "n"
colnames(dfTab)[2] <- "censor_text"
dfTab$lab <- as.character(100 * dfTab$Freq / sum(dfTab$Freq))

ggplot(dfTab, aes(x=n, y=Freq, fill=censor_text)) + geom_bar(stat='identity') +
  scale_fill_brewer(palette = "Set3") + labs(title = 'CD4 Frequency', x = "CD4", y = "Count") +  theme(legend.position = "right") +  theme(legend.title=element_blank()) + geom_text(aes(x=dfTab$n, y=dfTab$Freq, label=dfTab$Freq), position = position_stack(vjust = 0.5))

chisq.test(aids$censor_text, aids$cd4f)
```
The p-value of 9.129e-08 indicates that there is a significant difference in the proportion of individuals who develop AIDS or die across the different cd4 factor groups.

```{r, echo=FALSE}
ggplot(aids %>% filter(txgrp == 1 & censor == 1), aes(x=time)) + geom_histogram(breaks=seq(10, 80, by=5), col='black', fill='turquoise') + 
  labs(title="Histogram for Control Time to AIDS/Death", x="Time to AIDS/Death", y="Count") + 
  xlim(c(0,90)) + ylim(c(0,8))

ggplot(aids %>% filter(txgrp == 2 & censor == 1), aes(x=time)) + geom_histogram(breaks=seq(10, 80, by=5), col='black', fill='turquoise') + 
  labs(title="Histogram for IDV Time to AIDS/Death", x="Time to AIDS/Death", y="Count") + 
  xlim(c(0,90)) + ylim(c(0,8))
```

CD4 Stratification and Treatment Group:
```{r, echo=FALSE}
table(aids$txgrp_text, aids$strat2_text)
chisq.test(aids$txgrp_text, aids$strat2_text)
```
Subjects were eligible for the trial if their CD4 < 200 and had at least 3 months of prior zidovudine therapy. Randomization was stratified by CD4 cell count at the time of screening into groups of CD4 > 50 and CD4 <= 50. 50.8% of the CD4 <= 50 group recieved IDV while 49.8% of the CD4 > 50 group recieved IDV. The p-value of the chi-squared test of proportions  is very large (0.85), so there is no significant difference in the IDV and treatment proportions across the two CD4 groups.



CD4 Stratification and AIDS/Death Outcome: 
```{r, echo=FALSE}
table(aids$censor_text, aids$strat2_text)
chisq.test(aids$censor_text, aids$strat2_text)
```
4.0% of the CD4 <= 50 group and  14.7% of the CD4 > 50 group experienced AIDS/death. The p-value of the chi-squared test of proportions  is very small (6.015e-08), so there is a significant difference in the proportion of individuals that develop AIDS or die across the two CD4 groups.



Treatment Group and AIDS/Death Outcome:
```{r, echo=FALSE}
table(aids$censor_text, aids$txgrp_text)
chisq.test(aids$censor_text, aids$txgrp_text)
```
5.4% of the IDV group and 10.9% of the Placebo group experienced AIDS/Death. The p-value of the chi-squared test of proportions  is  small (0.004594), so there is a significant difference in the proportion of individuals that develop AIDS or die across the two treatment groups.


Treatment Group and Death: 
```{r, echo=FALSE}
table(aids$censor_d_text, aids$txgrp_text)
chisq.test(aids$censor_d_text, aids$txgrp_text)
```
1.4% of the IDV group and 3.3% of the Placebo group experienced death. The p-value of the chi-squared test of proportions  is  0.105, so we cannot conclude that there is a significant difference in the proportion of individuals that develop AIDS or die across the two treatment groups.


IV Usage and AIDS/Death:
```{r, echo=FALSE}
table(aids$censor_text, aids$ivdrug_text)
chisq.test(aids$censor_text, aids$ivdrug_text)
```
The p-value of 0.7401 indicates that there is no significant difference in the proportion of individuals who develop AIDS or die across the different drug usage groups. 


Hemophilia and AIDS/Death: 
```{r, echo=FALSE}
table(aids$censor_text, aids$hemophil_text)
chisq.test(aids$censor_text, aids$hemophil_text)
```
The p-value of 0.918 indicates that there is no significant difference in the proportion of individuals who develop AIDS or die across the two hemophilia groups. 
