---
title: "ExploratoryAnalysis"
author: "Amy Watt"
date: "2019-03-28"
output: html_document: 
  keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
aids <- read.csv("AIDSdata.csv")
library(tidyverse)
library(ggplot2)
```



```{r pressure, echo=FALSE}
ggplot(aids, aes(x=as.factor(sex), fill=as.factor(sex))) + geom_bar( ) +
  scale_fill_brewer(palette = "Set2") + labs(x = "Sex", y = "Count")

ggplot(aids, aes(x=as.factor(raceth), fill=as.factor(raceth))) + geom_bar( ) +
  scale_fill_brewer(palette = "Set2") + labs(x = "Race/Ethnicity", y = "Count")
```
```{r}
ggplot(aids, aes(x=age)) + geom_histogram(breaks=seq(10, 80, by=5), col='black', fill='turquoise') + 
  labs(title="Histogram for Age", x="Age", y="Count") + 
  xlim(c(10,80)) + ylim(c(0,450))
```
Adding a new column in dataframe for combined sex/raceth demographics.
11 = male, white non-hispanic
12 = male, black non-hispanic
13 = male, hispanic
14 = male, asian/pacific islander
15 = male, american indian/alaska native
16 = male, other/unknown
21 = female, white non-hispanic
22 = female, black non-hispanic
23 = female, hispanic
24 = female, asian/pacific islander
25 = female, american indian/alaska native
26 = female, other/unknown
```{r}
aids$demo <- NA
for (elt in 0:length(aids$id)) {
  s = toString(aids$sex[elt])
  r = toString(aids$raceth[elt])
  #aids$demo[elt] <- as.numeric(paste (s, r, sep=''))
  aids$demo[elt] <- (paste (s, r, sep=''))
}
```

```{r}
ggplot(aids, aes(x=as.factor(demo), fill=as.factor(demo))) + geom_bar( ) +
  scale_fill_brewer(palette = "Set3") + labs(x = "Demographics", y = "Count")
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
